---
title: Getting Started with unconfoundedr
author: Diogo Ribeiro
date: '`r Sys.Date()`'
output:
  'rmarkdown::html_vignette':
    toc: true
    toc_depth: 3
vignette: >-
  %\VignetteIndexEntry{Getting Started with unconfoundedr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `unconfoundedr` package provides tools for testing the **unconfoundedness assumption** in observational causal inference by comparing treatment effect estimates from RCT-like and observational datasets. This vignette demonstrates the main functionality through practical examples.

## Why Test Unconfoundedness?

When making causal inferences from observational data, we assume that all confounders are measured and controlled for (**unconfoundedness** or **ignorability**). However, this assumption is untestable with observational data alone.

**Key Insight**: If we have access to both experimental (RCT) and observational data on the same treatment-outcome relationship, we can compare their estimates. Large differences may indicate unmeasured confounding in the observational analysis.

```{r setup}
library(unconfoundedr)
set.seed(42)
```

# Basic Usage

## Generate Example Data

Let's start by creating example datasets with known properties:

```{r generate_data}
# RCT data: randomized treatment, no unmeasured confounding
rct_data <- generate_rct_data(
  n = 500,
  p_covariates = 3,
  treatment_effect = 1.0,
  outcome_type = "continuous",
  seed = 123
)

# Observational data: confounded treatment assignment
obs_data <- generate_obs_data(
  n = 1000,
  p_covariates = 3,
  treatment_effect = 1.0,        # Same true effect
  confounding_strength = 0.4,    # But with confounding bias
  covariate_shift = TRUE,        # And population differences
  outcome_type = "continuous",
  seed = 456
)

# Examine the data structure
head(rct_data)
head(obs_data)

# Treatment prevalence comparison
cat("RCT treatment prevalence:", mean(rct_data$A), "\n")
cat("Observational treatment prevalence:", mean(obs_data$A), "\n")
```

## Basic Unconfoundedness Test

Now let's run the basic test:

```{r basic_test}
result <- unconfoundedness_test(
  data_rct = rct_data,
  data_obs = obs_data,
  formula = Y ~ A + X1 + X2 + X3,    # A must be first RHS term
  estimator = "aipw",                 # Doubly robust estimator
  effect_measure = "rd",              # Risk difference
  transport = "none",                 # No transportability adjustment yet
  inference_method = "bootstrap",     # Bootstrap inference
  B = 500,                            # Bootstrap replicates (reduced for speed)
  seed = 42
)

# Print results
print(result)
```

## Interpreting the Results

The output shows:

- **RCT effect**: The treatment effect estimated from the RCT-like data
- **Observational effect**: The treatment effect from observational data
- **Difference**: The key quantity testing unconfoundedness
- **P-value**: Tests H0: difference = 0 (unconfoundedness holds)
- **E-value**: Robustness to unmeasured confounding

In this case, we expect to see a significant difference because we introduced confounding in the observational data.

# Advanced Features

## Transportability Weighting

When the RCT and observational populations differ, we can adjust for this using transportability weights:

```{r transportability}
# Test with automatic covariate shift detection
result_auto <- unconfoundedness_test(
  data_rct = rct_data,
  data_obs = obs_data,
  formula = Y ~ A + X1 + X2 + X3,
  estimator = "aipw",
  transport = "auto",          # Automatic shift detection
  auto_method = "both",        # Use both KS and energy tests
  auto_alpha = 0.01,           # Significance level for shift tests
  B = 500,
  seed = 42
)

print(result_auto)

# Check if transport was applied
cat("Transport applied:", result_auto$methods$transport_applied, "\n")
cat("Shift detected:", result_auto$diagnostics$auto_detection$shift_detected, "\n")
```

## Multiple Estimators

Compare results across different estimation methods:

```{r multiple_estimators}
estimators <- c("ipw", "aipw")  # Add "tmle", "gcomp" if packages available
results_by_estimator <- list()

for (est in estimators) {
  results_by_estimator[[est]] <- unconfoundedness_test(
    data_rct = rct_data,
    data_obs = obs_data,
    formula = Y ~ A + X1 + X2 + X3,
    estimator = est,
    transport = "auto",
    B = 300,  # Reduced for speed
    seed = 42
  )
}

# Compare estimates
comparison_df <- data.frame(
  Estimator = names(results_by_estimator),
  RCT_Effect = sapply(results_by_estimator, function(x) x$estimates$rct),
  Obs_Effect = sapply(results_by_estimator, function(x) x$estimates$observational),
  Difference = sapply(results_by_estimator, function(x) x$estimates$difference),
  P_Value = sapply(results_by_estimator, function(x) x$inference$p_value),
  stringsAsFactors = FALSE
)

print(comparison_df)
```

## Binary Outcomes and Different Effect Measures

For binary outcomes, we can use different effect measures:

```{r binary_outcomes}
# Generate binary outcome data
rct_binary <- generate_rct_data(
  n = 400,
  outcome_type = "binary",
  treatment_effect = 0.5,
  seed = 789
)
obs_binary <- generate_obs_data(
  n = 600,
  outcome_type = "binary",
  treatment_effect = 0.5,
  confounding_strength = 0.3,
  seed = 999
)

# Test different effect measures
effect_measures <- c("rd", "rr", "or")
binary_results <- list()

for (measure in effect_measures) {
  binary_results[[measure]] <- unconfoundedness_test(
    data_rct = rct_binary,
    data_obs = obs_binary,
    formula = Y ~ A + X1 + X2,
    effect_measure = measure,
    family_y = "binomial",
    estimator = "aipw",
    B = 300,
    seed = 42
  )
}

# Summary
binary_summary <- data.frame(
  Effect_Measure = names(binary_results),
  RCT_Effect = sapply(binary_results, function(x) x$estimates$rct),
  Obs_Effect = sapply(binary_results, function(x) x$estimates$observational),
  P_Value = sapply(binary_results, function(x) x$inference$p_value),
  stringsAsFactors = FALSE
)

print(binary_summary)
```

# Diagnostic Visualizations

The package provides several diagnostic plots:

```{r diagnostic_plots, fig.width=8, fig.height=6}
# Effect comparison plot
plot(result_auto, type = "effects")

# Propensity score overlap
plot(result_auto, type = "overlap")

# Covariate balance
plot(result_auto, type = "balance")

# Transport weights (if applied)
if (result_auto$methods$transport_applied) {
  plot(result_auto, type = "weights")
}

# Bootstrap distribution
plot(result_auto, type = "bootstrap")
```

# Detailed Diagnostics

## Summary Method

Get comprehensive diagnostics:

```{r detailed_summary}
# Comprehensive summary
summary_result <- summary(result_auto)
print(summary_result)
```

## Key Diagnostic Indicators

```{r diagnostic_indicators}
# Overlap quality
cat("Overlap quality:", result_auto$diagnostics$overlap$overlap_quality, "\n")

# Extreme propensity scores
cat("RCT extreme PS:",
    round(100 * result_auto$diagnostics$overlap$extreme_ps_proportion$rct, 1), "%\n")
cat("Obs extreme PS:",
    round(100 * result_auto$diagnostics$overlap$extreme_ps_proportion$obs, 1), "%\n")

# Sample sizes
cat("Effective sample sizes:\n")
cat("  RCT:", round(result_auto$diagnostics$sample_size$effective_n$rct), "\n")
cat("  Obs:", round(result_auto$diagnostics$sample_size$effective_n$obs), "\n")

# E-value interpretation
if (!is.null(result_auto$sensitivity$e_value)) {
  cat("E-value:", round(result_auto$sensitivity$e_value$e_value, 2),
      "(", result_auto$sensitivity$e_value$interpretation, ")\n")
}
```

# Power Analysis

Evaluate the power of the test under different scenarios:

```{r power_analysis}
# Small power analysis (use larger n_sim in practice)
power_result <- simulate_power_analysis(
  n_sim = 50,                       # Use 1000+ in practice
  n_rct = 300,
  n_obs = 500,
  treatment_effect_rct = 1.0,
  treatment_effect_obs = 1.3,       # 30% bias
  confounding_strength = 0.4,
  alpha = 0.05,
  seed = 123
)

print(power_result$results)

# Plot power simulation results
hist(power_result$raw_results$differences,
     main = "Distribution of Estimated Differences",
     xlab = "Difference (Obs - RCT)",
     col = "lightblue", border = "black")
abline(v = power_result$settings$true_difference, col = "red", lwd = 2, lty = 2)
legend("topright", "True Difference", col = "red", lty = 2, lwd = 2)
```

# Export and Reporting

## Export Results

```{r export_results, eval=FALSE}
# Export to CSV
export_to_csv(result_auto, file = "unconfoundedness_results.csv")

# Create LaTeX table
create_latex_table(result_auto, file = "results_table.tex",
                   caption = "Unconfoundedness Test Results")

# Generate diagnostic report
create_diagnostic_report(result_auto, file = "diagnostics_report.html",
                         format = "html")
```

## Batch Analysis

Process multiple dataset comparisons:

```{r batch_analysis}
# Create multiple dataset pairs
rct_studies <- list(
  study1 = rct_data[1:200, ],
  study2 = rct_data[201:400, ]
)

obs_studies <- list(
  study1 = obs_data[1:400, ],
  study2 = obs_data[401:800, ]
)

# Batch processing
batch_results <- batch_analysis(
  rct_data_list = rct_studies,
  obs_data_list = obs_studies,
  formula = Y ~ A + X1 + X2 + X3,
  estimator = "aipw",
  transport = "auto",
  B = 200,               # Reduced for speed
  validate = FALSE        # Skip validation for demo
)

print(batch_results)
```

# Practical Guidelines

## Choosing Parameters

### Estimator Selection
- **AIPW**: Best default choice (doubly robust)
- **IPW**: When outcome model is uncertain
- **TMLE**: When SuperLearner is available
- **G-computation**: When propensity model is uncertain

### Transport Method
- **"none"**: When populations are similar
- **"rct_to_obs"**: When populations definitely differ
- **"auto"**: Let the data decide (recommended)

### Sample Size Considerations

```{r sample_size}
# Estimate required sample sizes
required_n <- compute_required_sample_size(
  power = 0.8,
  alpha = 0.05,
  effect_size = 0.3,  # Expected difference
  ratio_obs_to_rct = 2
)
print(required_n)
```

## Interpretation Framework

| Scenario | P-value | E-value | Interpretation |
|----------|---------|---------|----------------|
| p > 0.05 | --      | > 2.0   | Supports unconfoundedness |
| p < 0.05 | < 1.5   | --      | Fragile, investigate |
| p < 0.01 | > 2.0   | --      | Significant but robust |
| p < 0.01 | < 1.5   | --      | Strong evidence of bias |

## Common Issues and Solutions

### Poor Overlap

```{r poor_overlap_demo, eval=FALSE}
# If overlap is poor, consider trimming
result_trimmed <- unconfoundedness_test(
  data_rct = rct_data,
  data_obs = obs_data,
  formula = Y ~ A + X1 + X2 + X3,
  trim = c(0.05, 0.95),     # Trim extreme 10%
  estimator = "aipw",
  B = 500
)
```

### Small Sample Sizes
- Increase bootstrap replicates (B)
- Use analytical inference when possible
- Consider combining studies (meta-analysis)

### Model Specification
- Include all relevant confounders
- Check model diagnostics
- Consider flexible models (splines, interactions)

# Validation and Testing

```{r validation}
# Validate package installation
validate_installation()

# Test with known null case (same data generation process)
rct_null <- generate_rct_data(n = 300, treatment_effect = 1.0, seed = 111)
obs_null <- generate_obs_data(n = 400, treatment_effect = 1.0,
                              confounding_strength = 0, seed = 222)  # No confounding

null_result <- unconfoundedness_test(
  rct_null, obs_null, Y ~ A + X1 + X2,
  estimator = "aipw", B = 300, seed = 333
)

cat("Null case p-value:", round(null_result$inference$p_value, 3), "\n")
cat("Should be > 0.05 for correct type I error control\n")
```

# Conclusion

The `unconfoundedr` package provides a comprehensive toolkit for assessing the credibility of observational causal inferences. Key takeaways:

1. **Always check transportability** - populations may differ between studies
2. **Use multiple estimators** - cross-validate your findings
3. **Examine diagnostics carefully** - overlap and balance are crucial
4. **Consider sensitivity analysis** - E-values provide robustness assessment
5. **Bootstrap for robust inference** - especially with small samples

For more advanced features and detailed methodology, see the package documentation and additional vignettes.

# Further Reading

- Hartman, E., & Hidalgo, F. D. (2018). An assessment of the augmented inverse propensity weighted estimator. _Political Analysis_, 26(2), 173-183.
- VanderWeele, T. J., & Ding, P. (2017). Sensitivity analysis in observational research: introducing the E-value. _Annals of Internal Medicine_, 167(4), 268-274.
- Stuart, E. A. (2010). Matching methods for causal inference: A review and a look forward. _Statistical Science_, 25(1), 1-21.

---

_This vignette demonstrates the main functionality of unconfoundedr. For questions or issues, please visit our [GitHub repository](https://github.com/DiogoRibeiro7/unconfoundedr)._
